---
title: "Exercise prediction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Data


```{r import}
trainURL <- url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
training <- read.csv(trainURL)
valURL <- url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
validation <- read.csv(valURL)
library(caret)
library(dplyr)
```

Removed index column and user_name, matching colname for class in training and problem id in testing
not the same - problem id used as placeholder for later class predictions 
```{r reformat}
training <- select(training, -X, -user_name)
validation <- select(validation, -X, -user_name)
validation <- rename(validation, classe = problem_id)
```

Create a training set and test set to test different models before using the validation set.   
```{r data partition}
inTrain <- createDataPartition(y = training$classe, p = 0.7, list = FALSE)
training <- training[inTrain, ]
testing <- training[ - inTrain, ]
```

In training set, most variables had 0 NA values, while several variables were NA for 13462 cases out of 13737 total.
As this left approximately 2% of the cases with non-NA values, these variables were dropped. 
(all variables were either 0 NAs or 13462, so this was done by keeping variables that had no NAs in the train set)
```{r na}
training <- training[ ,colSums(is.na(training))==0]
testing <- select(testing, one_of(colnames(training)))
validation <- select(validation, one_of(colnames(training)))
```

Decided to remove nearZeroVariance predictors to focus on predictors more likely to facilitate model fit.  Saved as 
seperate dataframe in case more predictors needed (starting with predictors that would not require transformations).

```{r nearZeroVar}
nzv <- nearZeroVar(training)
filterTraining <- training[ , -nzv]
filterTesting <- testing[ , -nzv]
filterValidation <- validation[ , -nzv]
```


##Set up parallel, train control
```{r parallel}
library(parallel)
library(doParallel)
mycluster <- makeCluster(detectCores() - 1) 
registerDoParallel(mycluster)
```

```{r train control}

fitControl <- trainControl(method = "cv",
                           number = 10,
                           allowParallel = TRUE)
```

##Trees
```{r rpart}
set.seed(457)
rpartModel <- train(classe~ ., data = filterTraining, method ="rpart", trControl = fitControl)
```


```{r rf}
set.seed(13337)
rfModel <- train(classe~ ., data = filterTraining, method ="rf", trControl = fitControl)

```
##LDA

```{r lda}
set.seed(42)
ldaModel <- train(classe ~ ., data = filterTraining, method = "lda", trcontrol = fitControl)
```


##Boosted Trees
```{r gbm}
set.seed(913)
gbmModel <- train(classe ~ ., data = as.data.frame(filterTraining), method = "gbm", verbose = FALSE)
```
##Predictions and accuracy tests

```{r predictions}
rpartPredict <- predict(rpartModel, filterTesting)
rfPredict <- predict(rfModel, filterTesting)
ldaPredict <- predict(ldaModel, filterTesting)
gbmPredict <- predict(gbmModel, filterTesting)

```

```{r accuracy}
confusionMatrix(rpartPredict, filterTesting$classe)
confusionMatrix(rfPredict, filterTesting$classe)
confusionMatrix(ldaPredict, filterTesting$classe)
confusionMatrix(gbmPredict, filterTesting$classe)
```